---
title: "Chapter 3: Fundamentals of Handling Time Series Data with R"
format: gfm
---

## 3.1 Object for Handling Time Series

* Base R's `ts` class is used for handling time series objects.

```{r}
Nile

Nile |> plot()
```

* Time series objects specify the start, end, and frequency of the data in the series. 

```{r}
co2 |> 
  ts(start = c(1959, 1),
     end = c(1997, 12),
     frequency = 12) |> 
  window(end = c(1965, 12)) |>
  plot()
```

```{r}
# demonstrate some of base R's ts plotting generics...

Nile |>
  
  # artificially shift Nile data back 5 years
  ts(start = 1866) |> 
  
  # join back with original unaltered Nile data
  ts.union(Nile)

# plot multiple time series together
(2 * Nile) |>
  cbind(Nile) |>
  ts.plot(lty = c("dashed", "solid"))
```

## 3.2 Handling of Time Information

* `tsp()` can extract or set a combination of start time, end time, and frequency from a `ts` object. 
* `time()` can extract a sequence of time values from the `ts` object.

```{r}
tsp(Nile)
time(Nile)
```

* `Date` class is good for holding time data with resolution to the day.
* `POSIXct/POSIXlt` is good for holding time data with resolution to the second.

>There are a whole host of tidy extensions to all of these object. I'm largely ignoring them for the sake of this book. 


